Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-05-20 paper/main.tex
last update for SAT 2023 submission!
SAT23 final: enhancements & typos

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use fewer states than equivalent automata with simpler
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA can be obtained for example by translating LTL
formulas with tools \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} or \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav} of the Owl library,
\texttt{ltl2tgba} of the Spot library~\cite{duret.16.atva2}, or
\texttt{ltl3tela}~\cite{major.19.atva}. There are also algorithms
processing these automata, for example the emptiness
check~\cite{baier.19.atva} or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

@@

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction method is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi,\rem}(T)\big).
\]
In fact, this corresponds to the shape of the QBF formulas we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi,\rem}(T)$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in CNF or in a different shape.

@@

\end{itemize}
By $\vec{e},\vec{n},\vec{i},\vec{f}$ we denote the vectors of all
variables of the form $e_t$, $n_{t,k}$, $i_{c,k}$, and $f_{c,k}$,
respectively. The constructed QBF formulas have the form 
\[
  \Phi_{C,K}(\vec{n},\vec{i},\vec{f})~~=~~\forall \vec{e}~.~\mcycle(\vec{e})\implies\big(\msat_\varphi(\vec{e}) \iff \msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})\big),
\]
where $\forall\vec{e}$ denotes the sequence composed of $\forall e_t$
for all variables $e_t$ and the subformulas $\mcycle(\vec{e})$,
$\msat_\varphi(\vec{e})$, and
$\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$ are defined as
follows. The subformula $\mcycle(\vec{e})$ will be defined as the last
one and we present three versions of it.

@@

\subsection{Intermediate version $\mcycle_2(\vec{e})$}
The intermediate version says that $\Te$ is nonempty,
$\Te\subseteq\delta_S$ for some SCC $S$, and for each state $q\in Q$
it holds that $\Te$ contains a transition leading to $q$ if and only
if it contains a transition leading from $q$. Formally,
\[
%  \mcycle_2(\vec{e})=\bigvee_{t\in\delta}e_t~\wedge~
  \mcycle_2(\vec{e})=\mcycle_1(\vec{e})~\wedge~
  \bigwedge_{q\in Q}\Big(\bigvee_{t'\in\delta~\cap~Q\times\Sigma\times 2^M\times\{q\}}\hspace{-1cm}e_{t'}\hspace{1cm}\iff
  \bigvee_{t''\in\delta~\cap~\{q\}\times\Sigma\times 2^M\times Q}\hspace{-1cm}e_{t''}\hspace{6mm}\Big).
\]
This condition is satisfied by every cycle, but also by some sets of
transitions that are not cycles. Some examples of such sets are provided in Figure~\ref{fig:inter}.

@@

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if for each set of
  states $S\subseteq Q$ it holds that
  \begin{enumerate}[A.]
  \item\label{en:1} all transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$) or
  \item\label{en:2} all transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$) or
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transition leading from a state in $S$
    to a state outside $S$ and a transition leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not hold for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradition that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.%\qed
\end{proof}

@@

Finally, we mention several three simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

The third optimization extends $\Phi_{j,C,K}$ into the conjunction
\[
   \Phi_{j,C,K}~\wedge \bigwedge_{c\in\{1,2,\ldots,C\}}\bigwedge_{k\in N_K} (\neg
   i_{c,k}\vee\neg f_{c,k}).
\]
The added part says that any cube does not contain both $\Inf k$ and
$\Fin k$ for any $k$. A cube with both $\Inf k$ and $\Fin k$ would be
useless as it cannot be satisfied by any run.

@@

If some call of the function
$\mathit{satisfiable}(\Phi_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In the case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance marks considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
did not decide within the time limit. If $X$ is \texttt{T}, a longer
timeout may lead to further reductions. If the multi-level reduction
is used, the automaton name contains the information from all
levels. For example, the name `\texttt{L1\_5\_U L2\_3\_U L3\_1\_T}'
means that level 1 reduced the number of marks to 6 (reduction to 5 is
impossible on this level), level 2 reduced it to 4, and level 3 to 2
as the QBF solver did not finish in the time limit when trying to
reduce the number of marks to
1.%when trying the automaton may be potentially reduced further.

@@

%\todo{Upravit experimenty: pridat dalsi 2 sady automatu.}
We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Three of the considered
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the number of
fails, the number of produced automata with less than two marks, and
the number of automata with at least two marks. The last number is
typeset in bold as these automata are actually used for the
experimental evaluation of our reduction technique.

@@

To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator), we have applied all
single-level reductions and the multi-level reduction, always with
incremental approach and without reducing of the number of cubes. We
do not reduce the number of cubes as our primary aim is to reduce the
number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finish within the 30~seconds
limit. The timeout of the last QBF query implies that the automaton
may be potentially further reduced by the reduction with a longer time
limit.

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reductions and automata sets. The column
    \emph{reduction} shows the cumulative percentage of saved
    acceptance marks and \emph{time} reports the cumulative reduction
    time. The column \textit{solver timeout} indicates the number of
    instances when the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

@@

Figure~\ref{fig:quantile} presents the time spent by multi-level
reduction on individual automata of each automata sets. The chart
shows a pleasing finding that for every set, most automata are reduced
in under a second and the high cumulative running times are caused by 
a relatively small number of complicated automata.

>>>>

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use fewer states than equivalent automata with simpler
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA can be obtained for example by translating LTL
(linear temporal logic) formulas with tools \texttt{ltl2dela}
(known as Delag)~\cite{muller.17.gandalf} or \texttt{ltl2dgra}
(known as Rabinizer~4)~\cite{kretinsky.18.cav} of the Owl library,
\texttt{ltl2tgba} of the Spot library~\cite{duret.16.atva2}, or
\texttt{ltl3tela}~\cite{major.19.atva}. There are also algorithms
processing these automata, for example the emptiness
check~\cite{baier.19.atva} or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

By choosing an appropriate format of the acceptance condition, one can
easily represent many classical kinds of automata over infinite words,
e.g. a B\"uchi automaton is a TELA with $M = \{\tacc{1}{1}\}$,
$\varphi = \Inf\tacc{1}{1}$, and the acceptance mark \tacc{1}{1} being
placed on all accepting transitions.
Similarly, a Rabin automaton with $k$ acceptance
pairs has $M = \{\tacc{5}{1}, \tacc{3}{1'} \ldots, \tacc{5}{k}, \tacc{3}{k'}\}$
and $\varphi = (\Fin\tacc{5}{1} \land \Inf\tacc{3}{1'}) \lor \ldots
\lor (\Fin\tacc{5}{k} \land \Inf\tacc{3}{k'})$.
A state-based acceptance can be represented by TELA by placing the
corresponding acceptance marks on all transitions leaving the accepting
states.

@@

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction method is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi,\rem}(T)\big).
\]
In fact, this corresponds to the shape of the QBF formulas we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi,\rem}(T)$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in conjunctive normal form (CNF)
or in a different shape.

@@

\end{itemize}
By $\vec{e},\vec{n},\vec{i},\vec{f}$ we denote the vectors of all
variables of the form $e_t$, $n_{t,k}$, $i_{c,k}$, and $f_{c,k}$,
respectively. The constructed QBF formulas have the form 
\[
  \Phi_{C,K}(\vec{n},\vec{i},\vec{f})~~=~~\forall \vec{e}~.~\mcycle(\vec{e})\implies\big(\msat_\varphi(\vec{e}) \iff \msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})\big),
\]
where $\forall\vec{e}$ denotes the sequence composed of $\forall e_t$
for all variables $e_t$ and the subformulas $\mcycle(\vec{e})$,
$\msat_\varphi(\vec{e})$, and
$\msat_{C,K}(\vec{e},\vec{n},\vec{i},\vec{f})$ are defined as
follows. Three versions of the subformula $\mcycle(\vec{e})$ will
be presented.

@@

\subsection{Intermediate version $\mcycle_2(\vec{e})$}
The intermediate version says that $\Te$ is nonempty,
$\Te\subseteq\delta_S$ for some SCC $S$, and for each state $q\in Q$
it holds that $\Te$ contains a transition leading to $q$ if and only
if it contains a transition leaving $q$. Formally,
\[
%  \mcycle_2(\vec{e})=\bigvee_{t\in\delta}e_t~\wedge~
  \mcycle_2(\vec{e})=\mcycle_1(\vec{e})~\wedge~
  \bigwedge_{q\in Q}\Big(\bigvee_{t'\in\delta~\cap~Q\times\Sigma\times 2^M\times\{q\}}\hspace{-1cm}e_{t'}\hspace{1cm}\iff
  \bigvee_{t''\in\delta~\cap~\{q\}\times\Sigma\times 2^M\times Q}\hspace{-1cm}e_{t''}\hspace{6mm}\Big).
\]
This condition is satisfied by every cycle, but also by some sets of
transitions that are not cycles. Some examples of such sets are provided in Figure~\ref{fig:inter}.

@@

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if for each set of
  states $S\subseteq Q$ one of the following conditions hold.
  \begin{enumerate}[A.]
  \item\label{en:1} All transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$).
  \item\label{en:2} All transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$).
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transition leading from a state in $S$
    to a state outside $S$ and a transition leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not hold for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradiction that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.%\qed
\end{proof}

@@

Finally, we mention three simple optimizations of the formula
construction, which are always applied in the rest of the paper.

@@

The third optimization extends $\Phi_{j,C,K}$ into the conjunction
\[
   \Phi_{j,C,K}~\wedge \bigwedge_{c\in\{1,2,\ldots,C\}}\bigwedge_{k\in N_K} (\neg
   i_{c,k}\vee\neg f_{c,k}).
\]
The added part says that no cube contains both $\Inf k$ and $\Fin k$
for any $k$. A cube with both $\Inf k$ and $\Fin k$ would be useless
as it cannot be satisfied by any run.

@@

If some call of the function
$\mathit{satisfiable}(\Phi_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton, included in the
generated HOA, encodes the reason for
it. In the case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance marks considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
did not decide within the time limit. If $X$ is \texttt{T}, a longer
timeout may lead to further reductions. If the multi-level reduction
is used, the automaton name contains the information from all
levels. For example, the name `\texttt{L1\_5\_U L2\_3\_U L3\_1\_T}'
means that level 1 reduced the number of marks to 6 (reduction to 5 is
impossible on this level), level 2 reduced it to 4, and level 3 to 2
as the QBF solver did not finish in the time limit when trying to
reduce the number of marks to
1.%when trying the automaton may be potentially reduced further.

@@

%\todo{Upravit experimenty: pridat dalsi 2 sady automatu.}
We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature, using all available formula sets implemented in
\texttt{genltl} 2.10.4.
The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. We used the tool
\texttt{ltlfilt} of the Spot library to simplify the formulas and
remove duplicates. Altogether, we have generated 348 unique LTL
formulas. These have been translated to deterministic TELA by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}, and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Three of the considered
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance marks as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions and homepages of
used translators. For each translator, it also provides the number of
fails, the number of produced automata with less than two marks, and
the number of automata with at least two marks. The last number is
typeset in bold as these automata are actually used for the
experimental evaluation of our reduction technique.

@@

To each automata set (that is, the automata with at least two marks
produced by an individual LTL translator), we have applied all
single-level reductions and the multi-level reduction, always with
incremental approach and without reducing the number of cubes. We
do not reduce the number of cubes as our primary aim is to reduce the
number of acceptance marks.
%\todo{Vysvetlit, proc neredukujeme pocet cubes?}
The timeout for each QBF query was set to 30 seconds.
%
All reductions have been performed by the tool \texttt{telatko}
built with Spot library version 2.10.4 and Z3 version 4.8.15. The
experiments have been run on a computer with Intel$^\text{®}$ Core™ i7-8700
processor and 32 GB of memory running Ubuntu 20.04.4. We used the tool
\texttt{autcross} of the Spot library to get the statistics of the
reduced automata and the running times.

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
column \textit{solver timeout} shows the number of automata for which
the last query to QBF solver did not finish within the 30~seconds
limit. The timeout of the last QBF query implies that the automaton
may be potentially further reduced by the reduction with a longer time
limit. The reduction ratio drop in higher levels, e.g. between level~1
and level~2 for \texttt{ltl2dgra}, happens due to the QBF solver timeouts
occuring earlier as the formula being solved in higher level is more
complex. For \texttt{ltl2dela} and \texttt{ltl2dgra}, the automata sets
do not contain any automaton where level~2 or level~3 achieves a better
result than level~1. However, all levels contribute to the reductions in
the multilevel setting.

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reductions and automata sets. The column
    \emph{reduction} shows the cumulative percentage of saved
    acceptance marks and \emph{time} reports the cumulative reduction
    time. The column \textit{solver timeout} indicates the number of
    instances where the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{10pt}
\input{table-data.tex}
\end{table}

@@

Figure~\ref{fig:quantile} presents the time spent by multi-level
reduction on individual automata of each automata set. The chart
shows a pleasing finding that for every set, most automata are reduced
in under a second and the high cumulative running times are caused by 
a relatively small number of complicated automata.
